Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-09-16 document/ms.tex
Including some of Garrett comments, see Issue #40.  Also printed the PKU document as it is.

\begin{abstract}
Magnetic fields temporarily increase the opacity in stellar photospheres, impacting both their instantaneous stellar flux and the duration and efficiency of their convective energy transport during pre-main sequence stellar evolution.  The observational manifestation of local magnetic fields, starspots on young stars can cover a large fraction of the stellar surface, causing the incorrect assessment of both the effective temperature and the luminosity and therefore the placement of sources on HR diagrams.  The singleton weak-lined T-Tauri star LkCa 4 shows exceptionally large photometric modulation attributable to star spots, making it an ideal testbed for measuring starspot properties.  Infrared spectral features in high resolution IGRINS spectra indicate that 80\% of the surface is covered in $\teff \sim 2750$ K patches of cool photosphere at the epoch of the observation.  The rest of the $20\%$ of the stellar surface can be thought of as either ambient photosphere or a hot spot of $\teff\sim4100$ K.  The spectral energy distribution, color variability, rotational modulation, and low-resolution optical/IR spectroscopy-- when registered to the filling factors at the epochs of their observations--strengthen the case for a large starspot filling factor and temperature contrast through time.  The revised effective temperature and luminosity make LkCa 4 appear much lower mass and much younger when placed on a pre-main sequence HR diagram, compared to previous estimates based only on optical spectroscopy and not taking into account the large coverage fraction of cool spots.  The derived starspot properties explain the extreme $V-$band photometric amplitude of 0.5$-$0.8 magnitudes.  Geometric considerations of spot coverage suggest that other heavily spotted young stars have been misplaced on the observational HR diagram, leading to incorrect estimates for ages and masses and contributing, in part, to the apparent age spread observed towards young clusters.
\end{abstract}

@@

The stellar photosphere is characterized as two photospheric components with a starspot temperature $\teffb$ and ambient photosphere temperature $\teffa$, with scalar solid angular coverages $\Omega_b$ and $\Omega_a$, respectively.    Starspots (or spot groups) are assumed to be cooler than the ambient photosphere, and they share the same average intrinsic and extrinsic stellar parameters otherwise; that is $(\vsini, \logg, \Z, v_z)$ are identical for starspots and ambient photosphere.  The composite mixture model for observed flux density can be written down as:

\begin{eqnarray} \label{eqn:mix_M}
S_{\mathrm{mix}} = \Omega_a B(\teffa)  + \Omega_b B(\teffb)
\end{eqnarray}

@@

\begin{eqnarray} \label{eqn:fill_factor}
f_{\Omega} = \frac{\Omega_b}{\Omega_a + \Omega_b}
\end{eqnarray}

@@

The starspot model includes two more free parameters than the standard \iancze\ model, namely $\teffb$ and $f_{\Omega}$.  The addition of these two parameters makes the MCMC sampling much more correlated than it was before because the relative contribution of the starspot and ambient photosphere are nearly degenerate over small changes in effective temperature.  This challenge motivated the second important change to the spectral inference framework, which involves technical aspects of switching from sampling the nuisance and stellar parameters separately in a blocked Gibbs framework to ensemble sampling using \texttt{emcee} \citep{foreman13}.  The affine-invariant \texttt{emcee} ensemble sampler is more resilient to correlations among stellar temperatures and fill factor than the Metropolis-Hastings sampler used in Gibbs sampling.  The practical effect of this switch is that all 14 stellar and nuisance parameters are fit simultaneously in a single spectral order, making stellar parameter estimates $\vt_{o}$ unique for each spectral order $o$, whereas the \iancze\ strategy had the power to provide a single set of stellar parameters $\vt$ that was based on all $N_{\rm ord}$ spectral orders.  The $N_{\rm ord}$ sets of inferences on $\vt_{o}$ are combined by weighted averaging point estimates of results from reliable orders, which offers resilience to flagrant calibration artifacts or conspicuous model mismatches.  This process can be thought of as a coarse approximation to the much more sophisticated spectral line outlier rejection described in \iancze.  No attempt was made to downweight spectral-line outliers using local kernels.

@@

Full-spectrum fitting was performed for 48 of the 54 available IGRINS spectral orders as described in Section \ref{sec:methods} and Appendix \ref{methods-details}.  The MCMC samples generally had 5000 steps with 40 walkers.  The stellar parameter ranges were $\logg \in [3.5, 4.0]$ and $\Z \in [-0.5, 0.5]$; the effective temperature range was $\teffa, \teffb \in [2700, 4500]$.  $H-$band fits had normal distribution priors in place: solar metallicity to $\pm0.05$ dex, $logg=3.8\pm0.1$, and $\vsini=29\pm5$ km/s.  The MCMC chains all appeared to converge after about 1500 steps.

@@

The IGRINS spectrum demonstrates some features that are reproduced only in the hot photosphere model, or only in the cool photosphere model.  Figure \ref{fig:specPostageStamp} shows a selection of such spectral features for a range of fill factors.  The temperature-sensitivity for the whole-order spectral fit is shown as a function of fill factor in the rightmost panels.  The unshown portion of the spectral order dilutes the temperature-sensitivity of these distinctive features, but overall the fill factor correlates with effective temperature in order to reproduce the observed line-depth ratio.  In some cases, featureless cool photosphere spectra veil isolated spectral lines predicted in the hot photosphere models.  In other cases, the hot photosphere model veils sequences of shallow spectral features predicted in the cool photosphere model.  The cool photosphere models predict shallow features because line blanketing from multiple indistinct molecular bands overlap from rotationally broadening.  This line blanketing has probably hampered efforts to identify isolated spectral features suitable for line-depth-ratio analysis in the spectra of rotationally broadened young stars.  One such feature has been previously identified in the near-IR: the OH 1.563 $\mu$m line noted by \citet{oneal01} shows a clear pattern of 3 lines in our data, with the central line exceeding the depth of the adjacent two lines, although the \PHOENIX\ models predict a non-negligible hot star contribution to the middle line at our range of effective temperatures.  In comparison, the forward-modeling technique described in this work thrives in the presence of long sequences of indistinct yet predictably correlated spectral features.  Similar forward-modelling strategies have successfully identified patterns of metal lines in the line-blanketed spectra of low metallicity stars \citep{kirby11,kirby15}.  Figures \ref{fig:Hband3x7} and \ref{fig:Kband3x7} in the Appendix show 42 of the $H-$ and $K-$band spectra on a log scale with a single random composite model spectrum overplotted.

@@

{\bf For Somers:  do we want a paragraph with a more explicit comparison to Somers \& Pinnsoneault (2015) and Feiden (2016)?}

@@

The spectral inference method assumed only two temperature components present in the photosphere.  In reality, starspots are probably described by a range of effective temperatures like umbra, penumbra, and plages on the sun's surface.  A model possessing three unique temperatures and two unique fill factors might better approximate the minor constituents of the emergent spectrum than our two temperature model can, but would always fit the data better and could lead to the possibility of overfitting.  The fit-quality of the existing two-temperature fits seems fairly strong through much of $H-$band and many $K-$band orders, so the contribution of minor photospheric constituents must be relatively low.  Interestingly, $K-$band has several spectral orders that converge towards hybrid temperatures in the vicinity of $\sim3500$ K.  Most of these orders were flagged as erroneous due to strong metal line spectral outliers or uncorrected calibration artifacts.  But some orders just showed mediocre fits where no permutation of hot and cool two-temperature models could describe all of the variance in the spectrum.  These $K-$band orders could be examined for a penumbra-like temperature component.  We caution that systematic errors in the models would ultimately limit the interpretability of models with more than two temperature components.  Space-based high-precision planetary transit spectrophotometry could offer an avenue to measure the morphologies and temperature distributions of starspots for the subset of transiting exoplanet host stars or eclipsing binaries.

@@

\subsection{Interpretation of the large filling factor of cool starspots}
We have shown that LkCa 4 exposes between 67\% and 83\% cool photosphere towards our viewing direction in 2015-2016, the most recent season for which there is data.  In previous seasons, like Fall 2004 (\emph{c.f.} Figures \ref{fig:PhotPhase} and \ref{fig:vband_spot}), the filling factor of cool spots could have exceeded 90\%.  This exceptionally large fraction of surface area covered by spots defies the conventional wisdom that starspots are the minority constituent of the stellar surface.  The evidence for this derived fill factor has been shown in the SED, high-resolution spectral inference, direct spectral detection, and TiO variability modelling.  Perhaps the strongest evidence for the large fill factor of cool photosphere is the large amplitude of $V-$band modulation, which cannot be explained with small spots, even if an effective temperature of absolute zero is assigned to the cool regions.

@@

The reason for standardizing the flux is a matter of practicality: the number of PCA eigenspectra components in the spectral emulator scales steeply with the pixel-to-pixel variance of $f_{\lambda}(\{\vt_{\ast}\}^\textrm{grid})$, increasing the computational cost of spectral emulation.  The choice to standardize fluxes makes no difference for modeling a single photospheric component.  But for two-component photosphere models, the relative flux of the two model spectra needs to be accounted for to get an accurate estimate of the areal coverage fraction of the cool spots, $f_{\Omega} \equiv \Omega_b/(\Omega_a+\Omega_b)$.  So we scale the mixture model in the following way:

\begin{eqnarray} \label{eqn:norm_scaling}
f_{\lambda, \mathrm{mix}} &=& f_{\mathrm{a}} \bar f_{\lambda, \mathrm{a}} \times \Omega_a + f_{\mathrm{b}} \bar f_{\lambda, \mathrm{b}} \times \Omega_b \\
q &=& q(\vt_{\ast})\equiv \frac{f_{\mathrm{b}}(\teffb, ...)}{f_{\mathrm{a}}(\teffa, ...)} \\
f_{\lambda, \mathrm{mix}}^{\prime} &=& \bar f_{\lambda, \mathrm{a}} \times \Omega_a + q \bar f_{\lambda, \mathrm{b}} \times \Omega_b
\end{eqnarray}

@@

The mixture model is a linear operation with all the same stellar extrinsic parameters, so we can re-use all the same post-processed eigenspectra $\widetilde{\mathbf{\Xi}}$, mean spectrum $\widetilde{\xi}_\mu$, and variance spectrum $\widetilde{\xi}_\sigma$, with the tildes representing all post processing \emph{except} the $\Omega$ scaling.  We calulate \emph{two} sets of eigenspectra weights $\mathbf{w}_{\in (\mathrm{a}, \mathrm{b})}$, and their associated mean and covariances following the Appendix of \iancze, and yielding:

\begin{equation}
  \mathsf{M}_{\mathrm{mix}}^\prime = \Omega_a (\widetilde{\xi}_\mu + \mathbf{X} \mathbf{\mu}_{\mathbf{w}, \mathrm{a}}) + q \Omega_b (\widetilde{\xi}_\mu + \mathbf{X} \mathbf{\mu}_{\mathbf{w}, \mathrm{b}})
\end{equation}

\begin{equation}
  \mathsf{C}_{\mathrm{mix}}^\prime = \Omega_a^2 \mathbf{X} \mathbf{\Sigma}_\mathbf{w, \mathrm{a}} \mathbf{X}^T + q \Omega_b^2 \mathbf{X} \mathbf{\Sigma}_\mathbf{w, \mathrm{b}} \mathbf{X}^T + \mathsf{C}
  \label{eqn:modC}
\end{equation}

>>>>

\begin{abstract}
Magnetic fields temporarily increase the opacity in stellar photospheres, impacting both their instantaneous stellar flux and the duration and efficiency of their convective energy transport during pre-main sequence stellar evolution.  The observational manifestation of local magnetic fields, starspots on young stars can cover a large fraction of the stellar surface, causing the incorrect assessment of both the effective temperature and the luminosity and therefore the placement of sources on HR diagrams.  The singleton weak-lined T-Tauri star LkCa 4 shows exceptionally large photometric modulation attributable to star spots, making it an ideal testbed for measuring starspot properties.  Infrared spectral features in high resolution IGRINS spectra indicate that 80\% of the surface is covered in $\teffb \sim 2750$ K patches of cool photosphere at the epoch of the observation.  The rest of the $20\%$ of the stellar surface can be thought of as either ambient photosphere or a hot spot of $\teffa \sim4100$ K.  The spectral energy distribution, color variability, rotational modulation, and low-resolution optical/IR spectroscopy-- when registered to the filling factors at the epochs of their observations--strengthen the case for a large starspot filling factor and temperature contrast through time.  The revised effective temperature and luminosity make LkCa 4 appear much lower mass and much younger when placed on a pre-main sequence HR diagram, compared to previous estimates based only on optical spectroscopy and not taking into account the large coverage fraction of cool spots.  The derived starspot properties explain the extreme $V-$band photometric amplitude of 0.5$-$0.8 magnitudes.  Geometric considerations of spot coverage suggest that other heavily spotted young stars have been misplaced on the observational HR diagram, leading to incorrect estimates for ages and masses and contributing, in part, to the apparent age spread observed towards young clusters.
\end{abstract}

@@

The stellar photosphere is characterized as two photospheric components with a starspot temperature $\teffb$ and ambient photosphere temperature $\teffa$, with scalar solid angular coverages $\Omega_{\mathrm{spot}}$ and $\Omega_{\mathrm{amb}}$, respectively.    Starspots (or spot groups) are assumed to be cooler than the ambient photosphere, and they share the same average intrinsic and extrinsic stellar parameters otherwise; that is $(\vsini, \logg, \Z, v_z)$ are identical for starspots and ambient photosphere.  The composite mixture model for observed flux density can be written down as:

\begin{eqnarray} \label{eqn:mix_M}
S_{\mathrm{mix}} = \Omega_{\mathrm{amb}} B(\teffa)  + \Omega_{\mathrm{spot}} B(\teffb)
\end{eqnarray}

@@

\begin{eqnarray} \label{eqn:fill_factor}
f_{\Omega} = \frac{\Omega_{\mathrm{spot}}}{\Omega_{\mathrm{amb}} + \Omega_{\mathrm{spot}}}
\end{eqnarray}

@@

The starspot model includes two more free parameters than the standard \iancze\ model, namely $\teffb$ and $f_{\Omega}$.  The addition of these two parameters makes the MCMC sampling much more correlated than it was before because the relative contribution of the starspot and ambient photosphere are nearly degenerate over small changes in $\teffa$, $\teffb$, and $f_{\Omega}$.  This challenge motivated the second important change to the spectral inference framework, which involves technical aspects of switching from sampling the nuisance and stellar parameters separately in a blocked Gibbs framework to ensemble sampling using \texttt{emcee} \citep{foreman13}.  The affine-invariant \texttt{emcee} ensemble sampler is more resilient to correlations among stellar temperatures and fill factor than the Metropolis-Hastings sampler used in Gibbs sampling.  The practical effect of this switch is that all 14 stellar and nuisance parameters are fit simultaneously in a single spectral order, making stellar parameter estimates $\vt_{o}$ unique for each spectral order $o$, whereas the \iancze\ strategy had the power to provide a single set of stellar parameters $\vt$ that was based on all $N_{\rm ord}$ spectral orders.  The $N_{\rm ord}$ sets of inferences on $\vt_{o}$ are combined by weighted averaging point estimates of results from reliable orders, which offers resilience to flagrant calibration artifacts or conspicuous model mismatches.  This process can be thought of as a coarse approximation to the much more sophisticated spectral line outlier rejection described in \iancze.  No attempt was made to downweight spectral-line outliers using local kernels.

@@

Full-spectrum fitting was performed for 48 of the 54 available IGRINS spectral orders as described in Section \ref{sec:methods} and Appendix \ref{methods-details}.  The MCMC samples generally had 5000 steps with 40 walkers.  The stellar parameter ranges were $\logg \in [3.5, 4.0]$ and $\Z \in [-0.5, 0.5]$; the \PHOENIX\ model spectrum temperature range\footnote{The \PHOENIX\ model spectra are indexed by $\teff$, but this term carries a different meaning for spotted stars than non-spotted stars.} was $\teffa, \teffb \in [2700, 4500]$.  $H-$band fits had normal distribution priors in place: solar metallicity to $\pm0.05$ dex, $logg=3.8\pm0.1$, and $\vsini=29\pm5$ km/s.  The MCMC chains all appeared to converge after about 1500 steps.

@@

The IGRINS spectrum demonstrates some features that are reproduced only in the hot photosphere model, or only in the cool photosphere model.  Figure \ref{fig:specPostageStamp} shows a selection of six such spectral features for a range of plausible fill factors.  In some cases, featureless cool photosphere spectra veil isolated spectral lines predicted in the hot photosphere models.  In other cases, the hot photosphere model veils sequences of shallow spectral features predicted in the cool photosphere model.  The cool photosphere models predict shallow features because line blanketing from multiple indistinct molecular bands overlap from rotational broadening.  This line blanketing has probably hampered efforts to identify isolated spectral features suitable for line-depth-ratio analysis in the spectra of rotationally broadened young stars.  One such feature has been previously identified in the near-IR: the OH 1.563 $\mu$m line noted by \citet{oneal01} shows a clear pattern of 3 lines in our data, with the central line exceeding the depth of the adjacent two lines, although the \PHOENIX\ models predict a non-negligible hot star contribution to the middle line for our range of ambient and spot temperatures.  In comparison, the forward-modeling technique described in this work thrives in the presence of long sequences of indistinct yet predictably correlated spectral features.  Similar forward-modelling strategies have successfully identified patterns of metal lines in the line-blanketed spectra of low metallicity stars \citep{kirby11,kirby15}.  Figures \ref{fig:Hband3x7} and \ref{fig:Kband3x7} in the Appendix show 42 of the $H-$ and $K-$band spectra on a log scale with a single random composite model spectrum overplotted.

@@

The IGRINS and ESPaDOnS spectra indicate indicate $\vsini\sim$ 28 km~s$^{-1}$ and a rotation period of 3.375 days. Combining these numbers gives $R\sin{i} \sim 1.87 R_{\odot}$. Our HR diagram analysis gives $R \sim 2.3R_{\odot}$. These numbers are consistent for $\sin{i} \sim 0.813$, or an inclination of about 35$^{\circ}$ from edge on. These values show broad consistency between rotational properties, spectral fitting values, and our interpretation of a tilted star revealing a circumpolar region with large areal coverage of starspots.  While this is obviously not proof of anything, the fact that all our different angles agree nicely would be a fun tidbit to point out.

@@

The spectral inference method assumed only two temperature components present in the photosphere.  In reality, starspots are probably described by a range of spot temperatures like umbra, penumbra, and plages on the sun's surface.  A model possessing three unique temperatures and two unique fill factors might better approximate the minor constituents of the emergent spectrum than our two temperature model can, but would always fit the data better and could lead to the possibility of overfitting.  The fit-quality of the existing two-temperature fits seems fairly strong through much of $H-$band and many $K-$band orders, so the contribution of minor photospheric constituents must be relatively low.  Interestingly, $K-$band has several spectral orders that converge towards hybrid temperatures in the vicinity of $\sim3500$ K.  Most of these orders were flagged as erroneous due to strong metal line spectral outliers or uncorrected calibration artifacts.  But some orders just showed mediocre fits where no permutation of hot and cool two-temperature models could describe all of the variance in the spectrum.  These $K-$band orders could be examined for a penumbra-like temperature component.  We caution that systematic errors in the models would ultimately limit the interpretability of models with more than two temperature components.  Space-based high-precision planetary transit spectrophotometry could offer an avenue to measure the morphologies and temperature distributions of starspots for the subset of transiting exoplanet host stars or eclipsing binaries.

@@

\subsection{Interpretation of the large filling factor of cool starspots}
We have shown that LkCa 4 exposes between 67\% and 83\% cool photosphere towards our viewing direction in 2015-2016, the most recent season for which there is data.  In previous seasons, like Fall 2004 (\emph{c.f.} Figures \ref{fig:PhotPhase} and \ref{fig:vband_spot}), the filling factor of cool spots could have exceeded 90\%.  This exceptionally large fraction of surface area covered by spots defies the conventional wisdom that starspots are the minority constituent of the stellar surface.  The evidence for this derived fill factor has been shown in the SED, high-resolution spectral inference, direct spectral detection, and TiO variability modelling.  Perhaps the strongest evidence for the large fill factor of cool photosphere is the large amplitude of $V-$band modulation, which cannot be explained with small spots, even if a spot temperature of absolute zero is assigned to the cool regions.

@@

The reason for standardizing the flux is a matter of practicality: the number of PCA eigenspectra components in the spectral emulator scales steeply with the pixel-to-pixel variance of $f_{\lambda}(\{\vt_{\ast}\}^\textrm{grid})$, increasing the computational cost of spectral emulation.  The choice to standardize fluxes makes no difference for modeling a single photospheric component.  But for two-component photosphere models, the relative flux of the two model spectra needs to be accounted for to get an accurate estimate of the areal coverage fraction of the cool spots, $f_{\Omega} \equiv \Omega_{\mathrm{spot}}/(\Omega_{\mathrm{amb}}+\Omega_{\mathrm{spot}})$.  So we scale the mixture model in the following way:

\begin{eqnarray} \label{eqn:norm_scaling}
f_{\lambda, \mathrm{mix}} &=& f_{\mathrm{amb}} \bar f_{\lambda, \mathrm{amb}} \times \Omega_{\mathrm{amb}} + f_{\mathrm{spot}} \bar f_{\lambda, \mathrm{spot}} \times \Omega_{\mathrm{spot}} \\
q &=& q(\vt_{\ast})\equiv \frac{f_{\mathrm{spot}}(\teffb, ...)}{f_{\mathrm{amb}}(\teffa, ...)} \\
f_{\lambda, \mathrm{mix}}^{\prime} &=& \bar f_{\lambda, \mathrm{amb}} \times \Omega_{\mathrm{amb}} + q \bar f_{\lambda, \mathrm{spot}} \times \Omega_{\mathrm{spot}}
\end{eqnarray}

@@

The mixture model is a linear operation with all the same stellar extrinsic parameters, so we can re-use all the same post-processed eigenspectra $\widetilde{\mathbf{\Xi}}$, mean spectrum $\widetilde{\xi}_\mu$, and variance spectrum $\widetilde{\xi}_\sigma$, with the tildes representing all post processing \emph{except} the $\Omega$ scaling.  We calulate \emph{two} sets of eigenspectra weights $\mathbf{w}_{\in (\mathrm{amb}, \mathrm{spot})}$, and their associated mean and covariances following the Appendix of \iancze, and yielding:

\begin{equation}
  \mathsf{M}_{\mathrm{mix}}^\prime = \Omega_{\mathrm{amb}} (\widetilde{\xi}_\mu + \mathbf{X} \mathbf{\mu}_{\mathbf{w}, \mathrm{amb}}) + q \Omega_{\mathrm{spot}} (\widetilde{\xi}_\mu + \mathbf{X} \mathbf{\mu}_{\mathbf{w}, \mathrm{spot}})
\end{equation}

\begin{equation}
  \mathsf{C}_{\mathrm{mix}}^\prime = \Omega_{\mathrm{amb}}^2 \mathbf{X} \mathbf{\Sigma}_\mathbf{w, \mathrm{amb}} \mathbf{X}^T + q \Omega_{\mathrm{spot}}^2 \mathbf{X} \mathbf{\Sigma}_\mathbf{w, \mathrm{spot}} \mathbf{X}^T + \mathsf{C}
  \label{eqn:modC}
\end{equation}
